Dynamic Discourse and Donkeys: Using Monoids for Language Semantics in Practice

### 2013-04-28 paper.tex
Before transfer to Visser syntax

\title{Musings in Dynamic Semantics for Donkeys}
\author{Thomas Dybdahl Ahle}
\date{March 9, 2013}

@@

\begin{abstract}
Your abstract goes here...

donkeys sentences
dynamic logic
disjunction, errors
monads
dependencies

From Donkeys to Predicates: Musings Into Dynamic Semantics and Logification
Donkification and Logification
Donkeys into Logic
From Farmers and Donkeys to Grammars and Predicates
Dependency ased
Discourse and Donkeys: A Dependency Based 
Discourse and Donkeys: A Monoid Based Language Grammar
Discourse and Donkeys: Monoid Based Language Semantics
Discourse and Donkeys: Using Monoids for Language Semantics
Dynamic Discourse and Donkeys: Using Monoids for Language Semantics

‘a brief description of what you did; no more than a page.’
The abstract provides a summary of the project report. It may contain some or all of the following moves, 
often, but not always, in the following order. Only use one or two sentences for each move. Keep the 
language and sentence structure simple. The reader should be able to read the abstract and obtain the 
necessary information quickly and efficiently.
Move 1:  Background to the project
Move 2: Purpose of the project
Move 3: Problem tackled
Move 4: Work carried out
Move 5: Results 
Move 6: Conclusions or implications
Move 7: Achievements of the project

Let's try it then

How to treat determinants/indeterminants
  1) ref
  2) quant
  3) vp -> sentence function

Ambiguity
  1) Strong/Weak

How to get a good logical representation
  1) Looking like language
  2) Compositionality

@@

To draw up this point, let's first look at a sentence that easily lets itself convert and then a couple that dont:
\begin{itemize}
\item `Socrates is a donkey': $donkey(Socrates)$. This is a classical logical example, treating Socrates as a constant. If we want to do without constants we can write: $\forall x(Socrates(x)\rightarrow donkey(x))$, though this does allow more than one Socrates to exist.
\item `All his donkeys are treated well': This sentence somewhat willingly let's itself translate into: $\forall y(his(x,y)\wedge donkey(y) \rightarrow treated\_well(y))$, which has a free variable $x$ refering to some person from earlier in the discourse.
\item `Every farmer beats his donkey': This sentence clearly has two possible meanings: $\exists y(his(x,y)\wedge donkey(y)\wedge \forall z(farmer(z) \rightarrow beats(z,y))$ if there is only one donkey, and $\forall x(farmer(x)\rightarrow \exists y (beats(x,y)\wedge his(x,y)\wedge donkey(y))$ if there is a donkey for each farmer.
\item `The happy donkey doesn't exist': Here we get into more problems. We might write: $\exists x(happy\_donkey(x)\rightarrow doesnt\_exist(x))$, but that asserts some sort of entity representing the non existing donkey. Alternatively we could write $\exists x(happy\_donkey(x)\rightarrow \bot)$, but that is another sentence: `There are no happy donkeys'.
\item `They donkeys are cold': This is an often used example of an `implied action sentence' and thought to have the same meaning as `Please close the stable door'. The best we can do here seems to be $please\_close\_the\_stable\_door()$ or perhaps $\exists x(stable\_door(x)\wedge please\_close(x))$.
\end{itemize}

In this paper we will work with a subset of English that translate reasonably well into logic. This unfortunately means that arguably the majority of sentences will be unaccesible to our methods. We will take a shot at ambigious sentences like the third one though.

%TODO: Perhaps this should be moved to the aknowledgements or something. It sets a very demoralizing tone during the rest of the paper.
Even more unfortunate than the problems above is it that we won't be able to do anything about the concerning welfare for donkeys around the world\cite{donkey2013sanctuary}, but must restrict ourselves to simply pointing it out. We will follow Geach\cite{geach1962reference} on this, and make `If a farmer owns a donkey, he beats it' our primary object of study.
%
\begin{align}
\exists x (farmer(x) \wedge \exists y (donkey(y)\wedge owns(x,y))) \rightarrow beats(x,y)
\end{align}
%
This is arguably the sentence that first comes to mind for a moderately trained logician. Translating an indeterminate `a farmer' into $\exists x(farmer(x))$ seems correct, but clearly the formula has problems with scope.
%
\begin{align}
\exists x (farmer(x) \wedge \exists y (donkey(y)\wedge owns(x,y) \rightarrow beats(x,y))) \label{wrong_donkey_2}
\end{align}
%
Moving the implication into the scope of $x$ and $y$ also doesn't work. Sentence \eqref{wrong_donkey_2} evaluates to true as long as a single farmer beats his donkey. In fact as long as a farmer exists it evaluates to true, since plugging the farmer into the $donkey$ predicate makes the anticedent false.
%
\begin{align}
\forall x (farmer(x) \rightarrow \forall y ( owns(x,y) \wedge donkey(y) \rightarrow beats(x,y))) \label{correct_donkey}
\end{align}
%
This turns out to be the correct formula. At least as long as you consider the use of `all' in natural language equal to the `all' in logic, which can be vacously true.

So what is going on? Most people think of determinates `the donkey' and indeterminates `a donkey' as being a reference to an existing donkey and an `introduction' of some new donkey into the conversation. However in a lot of sentences this turns out to not be the case.

How to treat determinants/indeterminants
  1) ref
  2) quant
  3) vp -> sentence function

Another problem with sentence \eqref{correct_donkey} is a lack of compositionality.

How to get a good logical representation
  1) Looking like language
  2) Compositionality

Finally let's have a look at the ambiguity we noted in the beginning of this chapter

Ambiguity
  1) Strong/Weak

If a farmer owns a donkey he beats it
\begin{equation}
\forall x . farmer(x) \rightarrow (\forall y . donkey(y) \wedge owns(x,y) \rightarrow beats(x,y))
\end{equation}

There are a lot of problems with this representation
\begin{itemize}
  \item It is not compositional. 
  dfdf
  \item It doesn't very well follow the order of the original sentence
  \item It is just not very easy to get right
\end{itemize}

More stuff from wikipedia:

Unfortunately, this translation leads to a serious problem of inconsistency. Indefinites must sometimes be interpreted as existential quantifiers, and other times as universal quantifiers, without any apparent regularity

Another reason for studying to so called `donkey sentences' is their particular use of referants.

The upshot of the foregoing observations is that, apparently, indefinites are neither quantifiers (dårlig konstruktion) nor referential (no farmer, etc.) terms, and this problem entrains another one, for as long as it unclear what indefinites mean, it will also remain obscure how they can serve as antecedents to pronouns. Later on we will see that very similar problems arise in seemingly disparate domains, like the interpretation of tense and presuppositions.

(1)  A definite is used to refer to something that is already familiar at the current stage 
of the conversation. An indefinite is used to introduce a new referent. 
This has been  labeled the `familiarity  theory of definiteness.'
When confronted with (l), the logically minded semanticist will notice immediately 
that it  presumes  something  very  objectionable:  that  definites and indefinites  are 
referring expressions.  For  only  if  there  is  a referent at  all  can there be  any question 
of its familiarity or novelty

In montague it is a function from verb to sentence

If there is a Chompsky universal grammar, we should be able to make a simple logic over it??

% Define block styles
\tikzstyle{decision} = [diamond, draw, fill=blue!20,
    text width=4.5em, text badly centered, node distance=2.5cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20,
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']

\begin{tikzpicture}[node distance = 3cm, auto]
    % Place nodes
    \node [block] (nl) {Natural Language};
    \node [block, right of=nl] (dr) {Dependency Representation};
    \node [block, right of=dr] (monad) {Enhanced DPL Representation};
    \node [block, right of=monad] (dpl) {Dynamic Predicate Logic};
    \node [block, right of=dpl] (prop) {First Order Logic};
    % Draw edges
    \path [line] (nl) -- (dr);
    \path [line] (dr) -- (monad);
    \path [line] (monad) -- (dpl);
    \path [line] [dotted] (dpl) -- (prop);
\end{tikzpicture}

% It is not possible
It is clear that not all sentences can be unambigiously translated from natural language to logic. A sentence like 	
\begin{quotation}
A farmer beats a donkey
\end{quotation}
Might be translated as 
\begin{equation}
\exists x ( farmer(x) \wedge \exists y ( donkey(y) \wedge beats(x,y)))
\end{equation}
Indicating that ``a farmer'' refers to some specific individual. Or we can might see it as a general rule, and translate it as
\begin{equation}
\forall x ( farmer(x) \rightarrow \exists y ( donkey(y) \wedge beats(x,y)))
\end{equation}
Indicating that every farmer has some donkey he beats. Or we could even read it as
\begin{equation}
\exists y ( donkey(y) \wedge \forall x ( farmer(x) \rightarrow beats(x,y)))
\end{equation}
Indicating that the poor donkey is common between all farmers.
% This seams wrong to me

In this paper we will not try to solve the above problem. (Though it could be interesting to generate multiple possibilities and perhaps use context to filter out the correct one). Instead we will try to show how some old method for doing this conversion can be enhanced with some new tools

More questions:

Should those three have the same meaning?
[1a]  Jane likes a painting.
[1b]  It is not the case that Jane doesn't like a painting.
[1c]  Jane likes a painting and either it is raining or it isn't.

direct translation not possible
`He saw the man with the telescope'
and even if we understand the syntax
`The students all attended one class'
has at least two meanings.
interpretation of discourse is influenced and guided by the common ground that exists between speaker and hearer

@@

Logicians and lingusts have come up with many different update logics since the 80s. I will quickly describe the four most cited ones, to give an idea of the landscape we're working in:
%
\begin{itemize}
\item Discourse Representation Theory, 1981 by Hans Kamp\cite{kamp1981theory}, is perhaps the best known approach to using update semantics for understanding language. Kamp takes a very philosophical approach, thinking of his logic as a model for how a person builds up a context during a conversation.

For example, say you you listen to `Some farmer owns no donkey. He beats it'. Discourse Representation Theory (DRT) builds this up from two blocks. Each consists of a set of referents and a set of properties using those. Underlined referents are referents that need to be `merged' with a properly `introduced' one:
%
\begin{align}
&[_1 x\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)] ] \nonumber\\
\oplus\ & [_1 \underline{v}, \underline{w}\mid beats(v,w) ] \nonumber
\end{align}
%
The rule for `$\oplus$' is simply to do a disjoint union of the referents and the properties:
%
\begin{align}
&[_1 x, \underline{v}, \underline{w}\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)], beats(v,w) ] \nonumber
\end{align}
%
At this step it is possible to perform anaphora related heuristics. Since $x$ is the first paramter of $farmer$, we assume it should be so for $beats$ as well:
%
\begin{align}
&[_1 x, \underline{v}, w\mid farmer(x), v=x, \neg[_2 y\mid donkey(y), owns(x,y)], beats(v,w) ] \nonumber\\
&[_1 x, w\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)], beats(x,w) ]\nonumber
\end{align}
%
On the other hand $y$ is not known in the scope of $beats$, so there is no way we can get rid of $w$. And we shouldn't, because in the sentence there is of course no way `it' can refer to `a donkey'.

@@

\item File Change Semantics, 1982 by Irene Heim\cite{heim1983file}, are based around the puzzle of discourse `referents' which may, as we have seen, sometimes act like quantifiers, sometimes referents and sometimes anticedents. Heim suggests using so called `file cards' which though developed independently, has a lot in common with DRT.

However, where DRT defines complex rules for merging different `discourse referents', in File Change Semantics (FCS) the part `added onto' a file gets to decide how the merge is made. Some how similar to how lambda functions in montague grammers give you a lot of flexibility. For example in case of `adding' $p$, an n-ary atomic predicate, to a file $F$:
%
\begin{align}
&Sat(F + p) = \{a \in Sat(F) \mid R(a_{i_1}, \dots, a_{i_n})\} \nonumber
\end{align}
%
Where $Sat$ is the function that takes a file to the set of `individuals' `satisfying' it.

@@

\tikzstyle{round} = [circle, draw, text width=3em, text centered]
\begin{tikzpicture}[->, >=stealth', semithick, node distance = 3cm, auto]
    % Place nodes
    \node [round] (beats) {beats};
    \node [round, below left of=beats] (x) {x};
    \node [round, left of=x] (farmer) {farmer};
    \node [round, below right of=beats] (z) {z};
    \node [round, right of=z] (donkey) {donkey};
    \node [round, below right of=x] (owns) {owns};
    \node [round, below left of=owns] (y) {y};
    % Draw edges
    \path (beats) edge node {agt} (x);
    \path (beats) edge node {obj} (z);
    \path (z) edge node {dependent} (x);
    \path (z) edge node {a} (donkey);
    \path (x) edge node {every} (farmer);
    \path (x) edge node {def-as} (y);
    \path (owns) edge node {agt} (y);
    \path (owns) edge node {obj} (z);
\end{tikzpicture}

@@

\item Dynamic Predicate Logic, 1991 by Groenendijk and Stokhof\cite{groenendijk1991dynamic}, is what I will focus on in the remains of this paper. In contrast to some of the previous models, the authors of Dynamic Predicate Logic (DPL) don't make any big philosophical claims on connections between their model and the brain. Instead they simply create a logic that is isomorphic to PL, while keeping a reasonable ammount of `update semantics' like properties, such as being closer syntactically to natural language.

DPL also doesn't provide any guidiance into resolving anaphora. Just like giving meaning to the actual concepts, this must be handled outside of the logic. We will see later in this paper that this stance makes sence, since the current state of the art anaphora resolvers are simply based on simple heuristics over noun phrases.

@@

The logic cleverly uses these advances in logical understanding to create an update logic in a pragmatic way, without any external constructs and a minimum of new syntax, while still getting all the benefits of compoitionality etc.

@@

(The parenthesis is only for clarity. The logic is 100\% associative). Notice how we introduce referants casually as we progress in the discourse. We even manage to look pretty similar to PL at the same time.

@@

The most interesting question is how we work with quantifiers. Which on earth relation could $\exists S$ be? We are going to take a hint from computer programming and consider the variable decleration: $S:=\bot$. We want to write this in DPL as $\exists S \cdot \neg S$, such that e.g. $P\overline{Q}[\exists S \cdot \neg S]P\overline{QS}$ is true. Now since $\cdot\neg S$ acts as a filter, removing all assignments that doesn't have $\overline{S}$, it must be that $\exists S$ is a `random reset, introducing $S$ and allowing it every value, such that $P\overline{Q}[\exists S]P\overline{QS}$ and $P\overline{Q}[\exists S]P\overline{Q}S$ are both true. The image of \eqref{dpl_all_states} under $\exists S$ has 16 possible states.

Let's have a look at the model theoretic definition of the semantics. Visser builds this up in two steps: First a relational algebra and then DPL on top of it. But since DPL is bascailly identical to the relational algebra, we are going to try and do it in one step.

@@

However let's first look at some of the implications of the above definitions. For example, how can we intuitively understand the DPL notion of $\rightarrow$? Well, first notice that many of the eqations start with $\alpha=\beta\wedge\dots$. These equations we can think of intuitively as having a closed scope. We saw earlier how DRT nicely determined that the `it' in `A farmer doesn't own a donkey. He beats it' couldn't refer to `a donkey' because the negation didn't let it's variable out. In DPL we get the same effect. Because an assignment must be the same before and after a negated clause, all new variables introduced inside the clause have disappeared.

The DPL $\psi_1\rightarrow\psi_2$ in particular can be seen as the filter, that if you can get `through' $\psi_1$ then you must also be able to get through $\psi_2$. But notice that $\gamma$ doesn't have to equal $\alpha$. Hence variables introduced in $\psi_1$ can well bind variables in $\psi_2$. This will show to be extreamly important in the treatment of donkey sentences later.

@@

Having defined seamingly very sound semantics for DPL above, we may start to notice certain oddities about our choice of using the relation union for `or'. An obvious alternative would have been the closed/static $\neg(\neg\psi_1\wedge\neg\psi_2)$ or $\neg\psi_1\rightarrow\psi_2$ considered by Groenendijk\cite{groenendijk1991dynamic}.

Actually there is an obvious definition of `or' that we have totally ignored in the previous chapter:
\begin{align}
\alpha[\psi_1\vee\psi_2]\beta
 :\equiv&\ \alpha[\neg(\neg\psi_1\wedge\neg\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha[\neg\psi_1\rightarrow\psi_2)]\beta \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha[\neg\psi_1]\gamma\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma((\alpha=\gamma\wedge\gamma\not\models\psi_1)\rightarrow\gamma\models[\psi_2]) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge \forall\gamma(\alpha=\gamma\rightarrow(\gamma\models\psi_1\vee\gamma\models\psi_2)) \nonumber\\
 \leftrightarrow&\ \alpha = \beta \wedge (\alpha\models\psi_1\vee\alpha\models\psi_2) \label{sem_or}
\end{align}

So does this fix our problem of binding between clauses? No. In fact this definition doesn't even allow binding from a clause and out.

a farmer doesn't own a horse or a donkey
\begin{align}
\exists x\cdot farmer(x) \cdot \neg(\exists y\cdot horse(y)\cdot owns(x,y) v \exists y\cdot donkey(y)\cdot owns(x,y))\nonumber\\
\exists x\cdot farmer(x) \cdot \neg\exists y\cdot horse(y)\cdot owns(x,y) \cdot \neg \exists y\cdot donkey(y)\cdot owns(x,y)\nonumber
\end{align}

>>>>

%Musings in Dynamic Semantics for Donkeys
%From Donkeys to Predicates: Musings Into Dynamic Semantics and Logification
%Donkification and Logification
%Donkeys into Logic
%From Farmers and Donkeys to Grammars and Predicates
%Dependency ased
%Discourse and Donkeys: A Dependency Based 
%Discourse and Donkeys: A Monoid Based Language Grammar
%Discourse and Donkeys: Monoid Based Language Semantics
%Discourse and Donkeys: Using Monoids for Language Semantics
%Dynamic Discourse and Donkeys: Using Monoids for Language Semantics in Practice

\title{Dynamic Discourse and Donkeys: Using Monoids for Language Semantics in Practice}
\author{Thomas Dybdahl Ahle}
\date{March 9, 2013}

@@

\begin{abstract}

%‘a brief description of what you did; no more than a page.’
%The abstract provides a summary of the project report. It may contain some or all of the following moves,  often, but not always, in the following order. Only use one or two sentences for each move. Keep the  language and sentence structure simple. The reader should be able to read the abstract and obtain the  necessary information quickly and efficiently.
%Move 1:  Background to the project
%Move 2: Purpose of the project
%Move 3: Problem tackled
%Move 4: Work carried out
%Move 5: Results 
%Move 6: Conclusions or implications
%Move 7: Achievements of the project

A donkey sentence is a sentence with a pronoun that is bound in semantics, but not in syntax. This make them difficult to capture in the scope of ordinary logic. In this paper we explore some of the ways that have been proposed to tackle this problem, in particular Albert Visser's e-monoid extension of Dynamic Predicate Logic, which allows us to write formulas deceivingly similarly looking to natural language.

We explain those concepts in depth, deriving or proving every result stated. We then start looking into implementing the logic on a computer, experimenting with different ways of defining the semantics. We further explore state of the art tools for dependency parsing and coreference resolution, the workings of which we also explain in depth.

Finally we put it all together in the Haskell programming language. We test out our machinery on a range of different sentences, showing all the converting steps from natural language to propositional logic. The problem is hard and the result is far from perfect, but it shows some interesting and less explored possibilities in computational linguistics.

@@

To draw up this point, let's first look at a sentence that easily lets itself convert and then a couple that don't:
\begin{itemize}
\item `Socrates is a donkey': $donkey(Socrates)$. This is a classical logical example, treating Socrates as a constant. If we want to do without constants we can write: $\forall x(Socrates(x)\rightarrow donkey(x))$, though this does allow more than one Socrates to exist.
\item `All his donkeys are treated well': This sentence somewhat willingly let's itself translate into: $\forall y(his(x,y)\wedge donkey(y) \rightarrow treated\_well(y))$, which has a free variable $x$ referring to some person from earlier in the discourse.
\item `Every farmer beats his donkey': This sentence clearly has two possible meanings: $\exists y(his(x,y)\wedge donkey(y)\wedge \forall z(farmer(z) \rightarrow beats(z,y))$ if there is only one donkey, and $\forall x(farmer(x)\rightarrow \exists y (beats(x,y)\wedge his(x,y)\wedge donkey(y))$ if there is a donkey for each farmer.
\item `The happy donkey doesn't exist': Here we get into more problems. We might write: $\exists x(happy\_donkey(x)\rightarrow doesn't\_exist(x))$, but that asserts some sort of entity representing the non existing donkey. Alternatively we could write $\exists x(happy\_donkey(x)\rightarrow \bot)$, but that is another sentence: `There are no happy donkeys'.
\item `They donkeys are cold': This is an often used example of an `implied action sentence' and thought to have the same meaning as `Please close the stable door'. The best we can do here seems to be $please\_close\_the\_stable\_door()$ or perhaps $\exists x(stable\_door(x)\wedge please\_close(x))$.
\item `Jane likes a painting': This is easy to translate to logic, but are we comfortable that it has the same meaning as the sentence `It is not the case that Jane doesn't like a painting' and even `Jane likes a painting and either it is raining or it isn't'? It is hard to imagine a language with strong logical foundations, where those three sentences wouldn't be equal.
\end{itemize}

In this paper we will work with a subset of English that translate reasonably well into logic. This unfortunately means that arguably the majority of sentences will be inaccessible to our methods. We will take a shot at ambiguous sentences like the third one though.

%TODO: Perhaps this should be moved to the aknowledgements or something. It sets a very demoralizing tone during the rest of the paper.
Even more unfortunate than the problems above is it that we won't be able to do anything about the concerning welfare for donkeys around the world\cite{donkey2013sanctuary}, but must restrict ourselves to simply pointing it out. We will follow Geach\cite{geach1962reference} on this, and make `If a farmer owns a donkey, he beats it' our primary object of study.
%
\begin{align}
\exists x (farmer(x) \wedge \exists y (donkey(y)\wedge owns(x,y)) \rightarrow beats(x,y))
\end{align}
%
This is arguably the formula that first comes to mind for a moderately trained logician, or simply what you expect by looking at the structure of the syntax tree. Translating an indeterminate `a farmer' into $\exists x(farmer(x))$ seems correct, but clearly the formula has problems with scope.
%
\begin{align}
\exists x (farmer(x) \wedge \exists y (donkey(y)\wedge owns(x,y) \rightarrow beats(x,y))) \label{wrong_donkey_2}
\end{align}
%
Moving the implication into the scope of $x$ also doesn't work. Sentence \eqref{wrong_donkey_2} evaluates to true as long as a single farmer beats his donkey. In fact as long as a farmer exists it evaluates to true, since plugging the farmer into the $donkey$ predicate makes the antecedent false.
%
\begin{align}
\forall x (farmer(x) \rightarrow \forall y ( owns(x,y) \wedge donkey(y) \rightarrow beats(x,y))) \label{correct_donkey}
\end{align}
%
This turns out to be the correct formula. At least as long as you consider the use of `all' in natural language equal to the `all' in logic, which can be vacuously true.

So what is going on? Most people think of determinates `the donkey' and indeterminates `a donkey' as being a reference to an existing donkey and an `introduction' of some new donkey into the conversation. However in a lot of sentences this turns out to not be the case. There doesn't seam to any simple way to translate them into logic. Even less does it seem intuitive how to use them as antecedents to pronouns.

The common way to deal with this problem is to make a big jump in power. Using a Montague Grammar we can define any noun phrase (NP) as a function from a verb phrase (VP) to a sentence. This way `if a farmer' might be a function $VP \to S \to S$, but even with this power, a working function is not easy to construct\cite{montague1973proper}.

Another problem with sentence \eqref{correct_donkey} is a lack of compositionality. This property which is vital if we want to build up a meaning of a conversation piece by piece. Say for example that our primary sentence was followed by the sentence `He also feeds it': $feed(x,y)$. How do we combine the first and the second sentence to give the meaning $\forall x(farmer(x)\rightarrow \forall y(owns(x,y)\wedge donkey(y)\rightarrow beat(x,y)\wedge feed(x,y)))$?

The point is that there is simply not a way to do this in PL. All variables in a formula are always closed off in their scope, and there are no logical operators that can `push' a new predicate into a closed scope. Our previous case might be a bit unfair, since we were trying to push something into a complicated implication, but even in case of the simple sentences `There is a farmer. He owns a donkey.' we wouldn't be able to build a meaning from the pieces.

One could argue that we can simply obtain compositionality using Montague functions like we discussed earlier (and will discuss more deeply in a later chapter), but what would be really nice is to have a simple logic that simply has these things built in. In this paper we are going to look into not only one, but a couple of such logics. In these logics we will be able to write things such as:
%
\begin{align}
(\exists x \cdot farmer(x)) \cdot (\exists y \cdot donkey \cdot owns(x,y))
\end{align}
%
Here the variables survive throughout the formula, which allows us to concatenate it with other formulas. The parenthesis are associative, so we don't even need them.
Later in the paper we will also look at a logic allowing us to write:
%
\begin{align}
(\bowtie \cdot \exists x \cdot Farmer(x) \cdot \exists y \cdot donkey(y) \cdot owns(x,y) \cdot \bowtie beats(x,y)) \cdot feeds(x,y)
\end{align}
%
Which allows us to be compositional even in the case of implications.

It is interesting to consider whether such logics can give us something on the way of translating speech to logic. In this paper we are going to look into exactly that, and we will combine it with the best available tools for dependency analysis and coreference resolution. This is the game plan:
\vspace{1em}
\tikzstyle{decision} = [diamond, draw, fill=blue!20,
    text width=4.5em, text badly centered, node distance=2.5cm, inner sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20,
    text width=5em, text centered, rounded corners, minimum height=4em]
\tikzstyle{line} = [draw, -latex']
\begin{tikzpicture}[node distance = 3cm, auto]
    % Place nodes
    \node [block] (nl) {Natural Language};
    \node [block, right of=nl] (dr) {Dependency Representation};
    \node [block, right of=dr] (monad) {Enhanced DPL Representation};
    \node [block, right of=monad] (dpl) {Dynamic Predicate Logic};
    \node [block, right of=dpl] (prop) {First Order Logic};
    % Draw edges
    \path [line] (nl) -- (dr);
    \path [line] (dr) -- (monad);
    \path [line] (monad) -- (dpl);
    \path [line] [dotted] (dpl) -- (prop);
\end{tikzpicture}

Going from step 2 to step 3 we are going to try and eliminate some of the ambiguity we saw earlier by outputting multiple possible sentences. In particular we will try and make a strong and a weak reading of every implication. However before we get into the details of how everything is implemented in code, we should look more into the nifty details of the theory applied.

@@

Logicians and linguists have come up with many different update logics since the 80s. I will quickly describe the four most cited ones, to give an idea of the landscape we're working in:
%
\begin{itemize}
\item Discourse Representation Theory, 1981 by Hans Kamp\cite{kamp1981theory}, is perhaps the best known approach to using update semantics for understanding language. Kamp takes a very philosophical approach, thinking of his logic as a model for how a person builds up a context during a conversation.

For example, say you you listen to `Some farmer owns no donkey. He beats it'. Discourse Representation Theory (DRT) builds this up from two blocks. Each consists of a set of referents and a set of properties using those. Underlined referents are referents that need to be `merged' with a properly `introduced' one:
%
\begin{align}
&[_1 x\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)] ] \nonumber\\
\oplus\ & [_1 \underline{v}, \underline{w}\mid beats(v,w) ] \nonumber
\end{align}
%
The rule for `$\oplus$' is simply to do a disjoint union of the referents and the properties:
%
\begin{align}
&[_1 x, \underline{v}, \underline{w}\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)], beats(v,w) ] \nonumber
\end{align}
%
At this step it is possible to perform anaphora related heuristics. Since $x$ is the first parameter of $farmer$, we assume it should be so for $beats$ as well:
%
\begin{align}
&[_1 x, \underline{v}, w\mid farmer(x), v=x, \neg[_2 y\mid donkey(y), owns(x,y)], beats(v,w) ] \nonumber\\
&[_1 x, w\mid farmer(x), \neg[_2 y\mid donkey(y), owns(x,y)], beats(x,w) ]\nonumber
\end{align}
%
On the other hand $y$ is not known in the scope of $beats$, so there is no way we can get rid of $w$. And we shouldn't, because in the sentence there is of course no way `it' can refer to `a donkey'.

@@

\item File Change Semantics, 1982 by Irene Heim\cite{heim1983file}, are based around the puzzle of discourse `referents' which may, as we have seen, sometimes act like quantifiers, sometimes referents and sometimes antecedents. Heim suggests using so called `file cards' which though developed independently, has a lot in common with DRT.

However, where DRT defines complex rules for merging different `discourse referents', in File Change Semantics (FCS) the part `added onto' a file gets to decide how the merge is made. Some how similar to how lambda functions in Montague grammars give you a lot of flexibility. For example in case of `adding' $p$, an n-ary atomic predicate, to a file $F$:
%
\begin{align}
&Sat(F + p) = \{a \in Sat(F) \mid R(a_{i_1}, \dots, a_{i_n})\} \nonumber
\end{align}
%
Where $Sat$ is the function that takes a file to the set of `individuals' `satisfying' it.

@@

\tikzstyle{round} = [circle, draw, text width=3em, text centered]
\begin{tikzpicture}[->, >=stealth', semithick, node distance = 3cm, auto]
    % Place nodes
    \node [round] (beats) {beats};
    \node [round, below left of=beats] (x) {x};
    \node [round, left of=x] (farmer) {farmer};
    \node [round, below right of=beats] (z) {z};
    \node [round, right of=z] (donkey) {donkey};
    \node [round, below of=x] (y) {y};
    \node [round, below of=z] (owns) {owns};
    % Draw edges
    \path (beats) edge node {agt} (x);
    \path (beats) edge node {obj} (z);
    \path (z) edge node {dependent} (x);
    \path (z) edge node {a} (donkey);
    \path (x) edge node {every} (farmer);
    \path (x) edge node {def-as} (y);
    \path (owns) edge node {agt} (y);
    \path (owns) edge node {obj} (z);
\end{tikzpicture}

@@

\item Dynamic Predicate Logic, 1991 by Groenendijk and Stokhof\cite{groenendijk1991dynamic}, is what I will focus on in the remains of this paper. In contrast to some of the previous models, the authors of Dynamic Predicate Logic (DPL) don't make any big philosophical claims on connections between their model and the brain. Instead they simply create a logic that is isomorphic to PL, while keeping a reasonable amount of `update semantics' like properties, such as being closer syntactically to natural language.

DPL also doesn't provide any guidance into resolving anaphora. Just like giving meaning to the actual concepts, this must be handled outside of the logic. We will see later in this paper that this stance makes sense, since the current state of the art anaphora resolvers are simply based on simple heuristics over noun phrases.

@@

The logic cleverly uses these advances in logical understanding to create an update logic in a pragmatic way, without any external constructs and a minimum of new syntax, while still getting all the benefits of compositionality etc.

@@

(The parenthesis is only for clarity. The logic is 100\% associative). Notice how we introduce referents casually as we progress in the discourse. We even manage to look pretty similar to PL at the same time.

@@

The most interesting question is how we work with quantifiers. Which on earth relation could $\exists S$ be? We are going to take a hint from computer programming and consider the variable declaration: $S:=\bot$. We want to write this in DPL as $\exists S \cdot \neg S$, such that e.g. $P\overline{Q}[\exists S \cdot \neg S]P\overline{QS}$ is true. Now since $\cdot\neg S$ acts as a filter, removing all assignments that doesn't have $\overline{S}$, it must be that $\exists S$ is a `random reset, introducing $S$ and allowing it every value, such that $P\overline{Q}[\exists S]P\overline{QS}$ and $P\overline{Q}[\exists S]P\overline{Q}S$ are both true. The image of \eqref{dpl_all_states} under $\exists S$ has 16 possible states.

Let's have a look at the model theoretic definition of the semantics. Visser builds this up in two steps: First a relational algebra and then DPL on top of it. But since DPL is basically identical to the relational algebra, we are going to try and do it in one step.

@@

However let's first look at some of the implications of the above definitions. For example, how can we intuitively understand the DPL notion of $\rightarrow$? Well, first notice that many of the equations start with $\alpha=\beta\wedge\dots$. These equations we can think of intuitively as having a closed scope. We saw earlier how DRT nicely determined that the `it' in `A farmer doesn't own a donkey. He beats it' couldn't refer to `a donkey' because the negation didn't let it's variable out. In DPL we get the same effect. Because an assignment must be the same before and after a negated clause, all new variables introduced inside the clause have disappeared.

Of course this `static negation' might not always be the right thing. Consider the sentences `It is not that case that a farmer doesn't own a donkey. He beats it'. This should be a good sentence, but we can't handle it. In the next chapter we will look at how this might be handled.

The DPL $\psi_1\rightarrow\psi_2$ in particular can be seen as the filter, that if you can get `through' $\psi_1$ then you must also be able to get through $\psi_2$. But notice that $\gamma$ doesn't have to equal $\alpha$. Hence variables introduced in $\psi_1$ can well bind variables in $\psi_2$. This will show to be extremely important in the treatment of donkey sentences later.

@@

A major problem with our use of `relation union' as disjunction is the way it allows seemingly sound sentences to combine into syntactically invalid ones. Consider the sentence `Either a farmer owns a donkey or he doesn't'. We would write that as
\begin{align}
\exists x\cdot farmer(x)\cdot(\exists y\cdot owns(x,y)\cdot donkey(y)\vee \bot) \label{owns_donkey_or_not}
\end{align}
Now concatenating this sentence with `he beats it': $beat(x,y)$ would give us a problem. In fact this problem would be more than a syntactic problem, because before we know the model and the existence of the donkey, we don't know if the sentence is valid or not.

Visser\cite{visser1999donkey} suggests that we should build our logic on sets of relations read disjunctively instead of just relations. This `possible world' scenario would transform \eqref{owns_donkey_or_not}$\cdot beats(x,y)$ into $\{\exists x\cdot farmer(x)\cdot\exists y\cdot donkey(y)\cdot owns(x,y)\cdot beats(x,y), \exists x\cdot farmer(x)\cdot\bot\cdot beats(x,y)\} = \{\exists x\cdot farmer(x)\cdot\exists y\cdot donkey(y)\cdot owns(x,y)\cdot beats(x,y), \bot\}$. This clearly is always well defined.

Visser's idea doesn't fully save us though. If we now consider `a farmer either owns a donkey or owns a horse':
\begin{align}
\exists x\cdot farmer(x)\cdot(&\exists y\cdot owns(x,y)\cdot donkey(y)\nonumber\\
                              &\vee \exists z\cdot owns(x,z)\cdot horse(z))
\end{align}
And combines that with `He beats the donkey': $beat(x,y)$, we get, in Visser's model, $\{\exists x\cdot farmer(x)\cdot\exists y\cdot owns(x,y)\cdot donkey(y)\cdot beats(x,y), \exists x\cdot farmer(x)\cdot\exists z\cdot owns(x,z)\cdot horse(z)\cdot beats(x,y)\}$. The second one is clearly not valid (though we wouldn't get a problem if there is no horse or farmer) which we might catch `at compile time' and replace it with something like `$\bot$' or `error'. This sort of error handling is discussed and implemented in the implementation section, but is not terribly interesting from a formal point of view.

Much more interesting are the following two questions: How could we have translated \label{owns_donkey_or_horse} so it would actually work? After all it seemed like a reasonable sentence until variable naming got us. And does the way variable binding works with our definition actually make sense?
